---
title: "Athaliana_guardcell_GRN"
output: html_document
date: "2025-06-28"
author: Christopher T Fazekas, Huang Lab @ NYU Biology 
data: Libault lab @ Mizzou, 
---

#Load Athaliana Dataset and prepare MINI-EX inputs

```{r}
library(Seurat)
library(pheatmap)
library(dplyr)
library(clusterProfiler)
library(org.At.tair.db)
library(dendextend)

#Change working directory 
setwd("/scratch/ctf277/guardGRN/Nguyen_et_al_2025")
```

```{r}
#load Athaliana dataset 
atha <- LoadSeuratRds("Athaliana_joinlayer_newannotation1.rds")
```

```{r}
#write out expression matrix
atha.expression.matrix <- as.data.frame(as.matrix(GetAssayData(object = atha, assay = "RNA", layer = "counts")))
write.table(atha.expression.matrix, "ATHA_INPUTS/ATHA_matrix.tsv", sep='\t', quote = FALSE)
```

```{r}
#calculate all cluster markers and write out 
Idents(atha) <- atha$seurat_clusters
ath_markers_all <- FindAllMarkers(atha, only.pos=TRUE)
write.table(ath_markers_all, "ATHA_INPUTS/ATHA_allMarkers.tsv", sep = "\t", quote=FALSE)
```

```{r}
#Cluster maps for seurat_cluster ids and annotated celltypes 
DimPlot(atha, reduction = "umap", group.by = "clusternameST")
DimPlot(atha, reduction = "umap", group.by = "seurat_clusters", label = TRUE)
```
##Based upon cell types and cluster identities above, annotate the ATHA_identities file (ATHA/ATHA_identities.tsv)
##of the format:
13	mesophyll 3
1	mesophyll 1
0	spongy mesophyll
10	phloem 2
2	mesophyll 2
4	epidermis/pavement 2
8	xylem
(...)

```{r}
#write out cluster annotation for each cell
cells2clusters <- FetchData(atha, vars = 'ident')
write.table(cells2clusters,"ATHA_cells2clusters.tsv", sep='\t',quote=FALSE,col.names = FALSE) 
```

##Run MINI-EX - config.log is "ATHA_NOMOTIF_OUTPUTS/ATHA_log.txt" - Arabidopsis TFs provided by MINI-EX package, primarily from TFDB (https://planttfdb.gao-lab.org/)
##NO MOTIF filtering - some genes of interest (such as FAMA) lack motif information and are filtered out otherwise 

##Network visualization with cytoscape - network.tsv file formed via make_networkfile.sh script 

```{r}
atha.regs <- read_tsv("ATHA_NOMOTIF_OUTPUTS/regulons/ATHA_rankedRegulons.tsv")
atha.regs.members <- read_tsv("ATHA_NOMOTIF_OUTPUTS/regulons/ATHA_regulons.tsv", col_names = FALSE)
#add colnames to members and merge with regulon info for a full regulon dataframe with all associated data and members, row-wise 
dim(atha.regs.members)
dim(atha.regs)
colnames(atha.regs.members) <- c("TF","cluster_id","members")
atha.regs <- atha.regs %>%
  mutate(cluster_id = paste0("Cluster_", sapply(strsplit(cluster, "_"), `[`, 3)))
atha.regs <- left_join(atha.regs,atha.regs.members, by = c("TF","cluster_id"))
dim(atha.regs)
atha.regs
```

```{r}
#filter for guard cell regulons and order by borda_clusterRank 
atha.regs.gc <- atha.regs %>% dplyr::filter(celltype == "guard cell") 
atha.regs.gc <- atha.regs.gc %>% arrange(., borda_clusterRank)
atha.regs.gc

#select members of guard cell regulons
regmem.atha <- atha.regs.gc %>% dplyr::select(TF,members)
regmem.atha
```

```{r}
#unlist members from regulons and form a list based on regulon id 
atha_gene_sets <- list()

unlist_mems <- function(regmem,gene_sets){
  for(i in 1:nrow(regmem)) {
  TF_name <- regmem$TF[i]
  genes <- unlist(strsplit(regmem$members[i], ","))
  gene_sets[[TF_name]] <- genes
  }
  return(gene_sets)
}

atha_gene_sets <- unlist_mems(regmem.atha,atha_gene_sets)
```

```{r}
#check TF expression by cell type to access specificity of identified modules 
pheat_TF_avg_celltype <- function(seurat_obj, gene_sets, name){
  validation_genes <- setdiff(rownames(seurat_obj), unlist(gene_sets))
  tf_expression <- FetchData(seurat_obj, vars = names(gene_sets), slot = "data", assay = 'RNA') 
  colnames(tf_expression) <- gsub("^rna_", "", colnames(tf_expression)) #remove assay data from colnames
  tf_avg <- aggregate(tf_expression, by = list(seurat_obj$cell_type), mean) #average by cell_type
  
  # Create the matrix and set column names to cell types
  heatmap_matrix <- t(as.matrix(tf_avg[,-1]))
  colnames(heatmap_matrix) <- tf_avg$Group.1  # Set cell type names as column names
  
  p <- pheatmap(
    heatmap_matrix,
    scale = "row",
    main = paste0(name, " Module TF Expression by cell type cluster"),
  )
  return(p)
}
pheat_TF_avg_celltype(atha, atha_gene_sets, "Arabidopsis")
```

```{r}
#Go enrichment for module genes 
runGO <- function(gene_sets, seurat_obj, database, keytype){
  
  #background of genes in seurat_obj 
  background_universe <- rownames(seurat_obj)

  go_results <- list()
  for(i in 1:length(gene_sets)) {
    genes_present <- gene_sets[[i]][gene_sets[[i]] %in% rownames(seurat_obj)]
    
    go_results[[names(gene_sets)[i]]] <- enrichGO(
      gene = genes_present,
      universe = background_universe,
      OrgDb = database,
      keyType = keytype,
      ont = "BP",
      pAdjustMethod = "BH",
      pvalueCutoff = 0.1,
      qvalueCutoff = 0.2,
      minGSSize = 5,
      maxGSSize = 500
    )
  }
  return(go_results)
}

#run atha for modules with expression background from atha 
atha_go <- runGO(atha_gene_sets, atha, org.At.tair.db, "TAIR")
```

```{r}
#after running enrichGO, filter noisy terms 
filter_go_results <- function(go_result) {
  if(is.null(go_result) || nrow(go_result@result) == 0) return(go_result)
  
  #remove terms (> 200 genes) or basic metabolic processes. terminal leaf development, removed "post-embryonic" and "capsule" - bacteria-related as well
  broad_terms <- c("metabolic process", "biosynthetic process", 
                   "glycogen", "carbohydrate metabolic", "catabolic process","post-embryonic", "capsule")
  
  filtered <- go_result@result[
    go_result@result$Count < 200 &  # Not too many genes
    !grepl(paste(broad_terms, collapse="|"), 
           go_result@result$Description, ignore.case=TRUE),
  ]
  
  go_result@result <- filtered
  return(go_result)
}

#apply filtering
atha_go_filtered <- lapply(atha_go, filter_go_results)
```


```{r}
#extract GO terms
go_term_lists <- list()

for(i in 1:length(atha_go_filtered)) {
  #check if the result exists and has terms 
  if(!is.null(atha_go_filtered[[i]]) && nrow(atha_go_filtered[[i]]@result) > 0) {
    #extract top 50 terms 
    go_term_lists[[names(atha_gene_sets)[i]]] <- atha_go_filtered[[i]]@result$Description[1:min(50, nrow(atha_go_filtered[[i]]@result))]
  }
}

#calculate Jaccard similarity
jaccard_matrix <- matrix(0, nrow = length(go_term_lists), ncol = length(go_term_lists))
rownames(jaccard_matrix) <- names(go_term_lists)
colnames(jaccard_matrix) <- names(go_term_lists)
for(i in 1:length(go_term_lists)) {
  for(j in 1:length(go_term_lists)) {
    if(length(go_term_lists[[i]]) > 0 && length(go_term_lists[[j]]) > 0) {
      intersection <- length(intersect(go_term_lists[[i]], go_term_lists[[j]]))
      union <- length(union(go_term_lists[[i]], go_term_lists[[j]]))
      jaccard_matrix[i,j] <- intersection / union
    }
  }
}

#convert Jaccard to distance
jaccard_dist <- as.dist(1 - jaccard_matrix)

#cluster 
hc <- hclust(jaccard_dist, method = "complete")

#create dendrogram with colors
dend <- as.dendrogram(hc)
dend <- color_branches(dend, k = 7)  # Color by 3 groups

plot(dend, main = "TF Module Clustering by Functional Similarity")
```

```{r}
# cluster modules into groups based on level in dendrogram 
k <- 7 
module_groups <- cutree(hc, k = k)

#find common and unique GO terms for each group 
for(group in 1:k) {
  modules_in_group <- names(module_groups)[module_groups == group]
  cat("\n=== Functional Group", group, "===\n")
  cat("Modules:", paste(modules_in_group, collapse = ", "), "\n")
  
  if(length(modules_in_group) > 1) {
    #find shared GO terms within group
    shared_terms <- Reduce(intersect, go_term_lists[modules_in_group])
    cat("Shared functions (", length(shared_terms), " terms):\n")
    if(length(shared_terms) > 0) {
      cat(paste("-", shared_terms, "\n"))
    }
  }
  
  #unique terms for each module in group
  for(mod in modules_in_group) {
    other_modules <- setdiff(names(go_term_lists), mod)
    other_terms <- unique(unlist(go_term_lists[other_modules]))
    unique_terms <- setdiff(go_term_lists[[mod]], other_terms)
    cat("\nUnique to", mod, "(", length(unique_terms), " terms):\n")
    if(length(unique_terms) > 0) {
      cat(paste("-", head(unique_terms, 10), "\n"))
    }
  }
}
```

```{r}
analyze_module_groups <- function(hc, go_term_lists, k = 7) {
  #cluster modules into groups
  module_groups <- cutree(hc, k = k)
  
  #initialize result dataframe
  result_df <- data.frame(
    module = character(),
    group_id = integer(),
    shared_functions = character(),
    unique_functions = character(),
    stringsAsFactors = FALSE
  )
  
  #process each group
  for(group in 1:k) {
    modules_in_group <- names(module_groups)[module_groups == group]
    
    #find shared GO terms within group if >1 module
    shared_terms <- character()
    if(length(modules_in_group) > 1) {
      shared_terms <- Reduce(intersect, go_term_lists[modules_in_group])
    }
    
    #convert shared terms to single string
    shared_functions_str <- if(length(shared_terms) > 0) {
      paste(shared_terms, collapse = "; ")
    } else {
      ""
    }
    
    #process each module in the group
    for(mod in modules_in_group) {
      #find unique terms for this module
      other_modules <- setdiff(names(go_term_lists), mod)
      other_terms <- unique(unlist(go_term_lists[other_modules]))
      unique_terms <- setdiff(go_term_lists[[mod]], other_terms)
      
      #convert unique terms to single string
      unique_functions_str <- if(length(unique_terms) > 0) {
        paste(unique_terms, collapse = "; ")
      } else {
        ""
      }
      
      # Add row to result dataframe
      result_df <- rbind(result_df, data.frame(
        module = mod,
        group_id = group,
        shared_functions = shared_functions_str,
        unique_functions = unique_functions_str,
        stringsAsFactors = FALSE
      ))
    }
  }
  
  return(result_df)
}

#create module analysis df
module_analysis_df <- analyze_module_groups(hc, go_term_lists, k = 7)
#write out 
write.table(module_analysis_df, file = "GO_modules.tsv", row.names = FALSE, quote = FALSE, sep ='\t')
```

```{r}
#check which modules have GO data
cat("Modules with GO data:", length(atha_go_filtered), "\n")
cat("Gene sets:", length(atha_gene_sets), "\n")

#check for mismatches
go_modules <- names(atha_go_filtered)
gene_set_modules <- names(atha_gene_sets)
cat("Missing in GO:", setdiff(gene_set_modules, go_modules), "\n")
cat("Missing in gene sets:", setdiff(go_modules, gene_set_modules), "\n")

#check what gets filtered out
modules_with_data <- c()
for(i in 1:length(atha_go_filtered)) {
  if(!is.null(atha_go_filtered[[i]]) && nrow(atha_go_filtered[[i]]) > 0) {
    modules_with_data <- c(modules_with_data, names(atha_gene_sets)[i])
  }
}
cat("Modules with GO terms:", length(modules_with_data), "\n")
cat("Missing modules:", setdiff(names(atha_gene_sets), modules_with_data), "\n")

#check merge results before and after merge
cat("Modules before merge:", length(unique(top_terms_per_module$Module)), "\n")
cat("Modules after merge:", length(unique(top_terms_per_module$Module)), "\n")
```

```{r}
#extract top GO terms for each module
top_terms_per_module <- data.frame()
for(i in 1:length(atha_go_filtered)) {
  if(!is.null(atha_go_filtered[[i]]) && nrow(atha_go_filtered[[i]]) > 0) {
    df <- as.data.frame(atha_go_filtered[[i]])
    df <- df[1:min(4, nrow(df)), ]  
    df$Module <- names(atha_gene_sets)[i]
    top_terms_per_module <- rbind(top_terms_per_module, df)
  }
}

#color mapping for module groups
group_colors <- c(
  "1" = "#377EB8",
  "2" = "#4EB3D3",  
  "3" = "#74C476",
  "4" = "#C994C7",  
  "5" = "#6A51A3",
  "6" = "#8C96C6",  
  "7" = "#FC8D59"   
)

#module group mapping
module_names <- names(atha_gene_sets)
module_group_df <- data.frame(
  Module = module_names,
  Group = module_groups[module_names]
)

#merge with the plotting data
top_terms_per_module <- merge(top_terms_per_module, module_group_df, by = "Module")

#order modules by their group number
module_order <- module_group_df[order(module_group_df$Group), "Module"]
top_terms_per_module$Module <- factor(top_terms_per_module$Module, levels = module_order)

#create colors for x-axis text based on module groups
axis_colors <- setNames(
  group_colors[as.character(module_group_df$Group[match(module_order, module_group_df$Module)])],
  module_order
)

#create the plot
dot_atha <- ggplot(top_terms_per_module, aes(x = Module, y = Description, 
                                  size = Count, color = p.adjust)) +
  geom_point() +
  scale_color_gradient(low = "red", high = "blue") +
  scale_size_area(max_size = 4) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.7, size = 7, hjust = 2,
                                   color = axis_colors[levels(top_terms_per_module$Module)]),
        axis.text.y = element_text(angle = -30, hjust = 1, vjust = 1, size = 7)) + 
  labs(title = " ",
       x = " ", 
       y = " ",
       size = "Gene Count", 
       color = "Adjusted P-value")

ggsave("GO_atha_guard.png", dot_atha, scale = 1, width = 8, height = 10)
```
